Semantic Novelty Detection with Graphical Models

### 2011-05-05 semantic-nd/article.tex
Several changes in both dora description and novelty detection.

\begin{figure}[h]
\centering
\caption{The conceptual layer structures the sensed environment together with the conceptual knowledge in order to
         create a structured probabilistic representation of the world.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
\begin{itemize}
\item Explain 
\item Approach with a threshold function
\item Need of a normalizing factor for a threshold with dynamic functions
\item Usage of conditional density
\item Assuming of every output is equally likely
\item Using unlabelled data
\item What happens if the conditional and unconditional graphs differ only by the addition of a node?
\item How to implement that in the case a graph is a tree
\item Novelty detection on any variable of the graphical model!
\end{itemize}

Novelty detection can be performed by a threshold function sorting the input space in an optimal way to
decrease error in detection.

Its harder than a classification problem as the system only has access to positive samples.
Under that a common approach is to use conditional probability $P(x|known)$ as a threshold.

\subsection{Stable threshold}
Although in our case we are dealing with a dynamic graph $G$ that keeps changing through time as new variables
are added. Which increases the dimension of our space and spreads the distribution over the variables.

This turns it impossible to directly train a threshold function on $P_G(x)$, as we cannot assume $P(x)$ to be constant over any set of $x$.
So we need to describe $P(x)$ to develop a stable threshold on $P_G(x)/P(x)$.

A common assumption is that every possible input is equally likely then the threshold
should be $P_G(x)/(\prod |x_i|)$.

\subsection{Unlabelled data}
Nonetheless in the case of robotics its often the case we can access large amounts of unlabelled data.
This leads to an interesting point that instead of assuming a constant $P(x)$ we can actually try to build
model for it.

Under that we decided to test assuming that all probabilities are independent. That can be modelled as a graph $G'$.
And corresponding function $P_G(x)/P_{G'}(x)$.

This way our unconditional function perceive which features are slightly more biased to a certain value and will
compensate on the probability calculation.

This step is important to achieve a correct order of the input space.

>>>>

\begin{figure}[h]
\centering

\includegraphics[width=0.50\textwidth]{figures/conceptual-layer.jpg}
\includegraphics[width=0.49\textwidth]{figures/chain-graph.png}
\caption{The conceptual layer structures the sensed environment together with the conceptual knowledge in order to
         create a structured probabilistic representation of the world.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
\begin{itemize}
\item Define novelty detection 
\item Explain approach with a threshold function
\item Need of a normalizing factor for a threshold with dynamic functions
\item Usage of conditional density
\item Assuming of every output is equally likely
\item Using unlabelled data
\item What happens if the conditional and unconditional graphs differ only by the addition of a node?
\item How to implement that in the case a graph is a tree
\item Novelty detection on any variable of the graphical model!
\end{itemize}

% Introduce novelty detection
Novelty detection is an harder problem than that of classification due to its single-class nature.
Essentially a system is given training data and when given new data has to decide if it belongs
to the same concept it was trained with.

% Explain threshold approach to novelty detection
Due to the desired of robustness and generalization, a novelty detection system would always possess
some threshold that describes on how strict the system should be to new data.
This threshold is often seen as a distance measure to a normal input and several novelty detection
techniques are based on it.

When dealing with a statistic point of view, noisy data or unstable features, a decision on a given input has
an associated error rate. And in this case a novelty detection system is interested in defining an order
on all the possible inputs equivalent to the order imposed by the error rate: $P(novel|x)$.

Such an ordering can be defined by a R-value function on the input space. And then a threshold can be trained
on that function in order to accept an expected amount of cases at the cost of failing in others.

Based on bayes rule:

\begin{equation}
P(\overline{novel}|x) = \frac{P(x|\overline{novel}) P(\overline{novel})}{P(x)}
\end{equation}

% Explain approach of using conditional probability
\subsection{Conditional Probability: $P(x|\overline{novel})$}
One of the hardest problems is that on most cases novelty detection system only have access to 
the known data.
Under that the best approach is to define a threshold based on $P(x|\overline{novel})/P(x)$
considering that $P(x)$ is constant through all the input space.

In the presented case, $P(x|\overline{novel})$ is approximated by the graphical model $G$ produced
by the conceptual layer. As so we use $P_G(x)$ as a substitute.

\label{sec:uniform-unconditional}
In several cases assuming the unconditional probability $P(x)$ is constant leads to discarding the factor,
nonetheless here the distributions are dynamically changing as the robot learns more on the environment.
So we need to keep the normalizing argument $P(x)$ and re-calculate it for each new subset of $x$.
Still assuming that the unconditional distributions generates all possible outcome with the same probability
we can model it $P(x)=\prod{|x_i|}$, where $|x_i|$ denotes the state space size of variable $x_i$.
In graphical model terms, this would be equivalent to build a factor graph $G'$ with the same variables but
without any factors.

Using a graphical model built to model the known data and assuming that in the unconditional case all possible outcomes
are equally likely a novelty threshold would be given by: $P_G(x)/P_{G'}(x)$, as illustrated on \autoref{fig:unsupervised-threshold}.

\begin{figure}[h]
\caption{\label{fig:unsupervised-threshold}By producing graphical models $G$ and $G'$ to both conditional and unconditional probability
         distributions on a set of variables, novelty can be detect on the threshold $P_G(x)/P_{G'}(x)$.
         In this case the unconditional probability is modelled as a fully uniform distribution.}
\end{figure}

\subsection{Using Unlabelled Data}
Nonetheless in several cases its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability distribution that
the uniform one assumed in \autoref{sec:uniform-unconditional}.

For interesting and practical reasons it was assumed that the unconditional distribution could be modelled with
all the variables independent from each other.
Which is the same as assuming $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate $P_{x_i}$.
Once again \autoref{fig:semisupervised-threshold} illustrates this in terms of graphs $G$ and $G'$.

\begin{figure}[h]
\caption{\label{fig:semisupervised-threshold}Using unlabelled data, its possible to create a better approximation
         for the unconditional probability. Which should lead to a better novelty threshold.}
\end{figure}

The described ratio between the two graphs can be understood as compensating for an existing bias
on the unconditional distribution.
Which is an important step to achieve a correct ordering of the input space for novelty thresholding.
This way our unconditional function perceive which features are slightly more biased to a certain value and will
compensate on the probability calculation.

\subsection{Generalizing to Novelty Detection on a variable}
Until now, the presented methods describe how to detect data that belongs to a new class of input by creating
generative models for both conditional and unconditional probabilities distributions.

And although it serves as a method for detecting new classes in the input, it gives us no control in what
variables novelty is being measured.

Lets assume that the only difference between the conditional and unconditional graph is a single node and its connecting factors
as seen on figure X.
